BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2017-03-30 Paper2/BOAST.tex
Modification of SPECFEM3D section. Answers point 4) and 5) of reviewer.

%%% Porting SPECFEM3D to OpenCL using BOAST %%%
\subsection{Porting SPECFEM3D to OpenCL using BOAST\label{subsec:specfem}}

In this last subsection, we study a free seismic wave propagation simulator,
SPECFEM3D, in its GLOBE flavor\footnote{Specfem3d Globe --- CIG
\url{http://www.geodynamics.org/cig/software/specfem3d-globe}}. SPECFEM3D
simulates seismic wave propagation at the local or regional scale based upon
spectral-element method (SEM), with very good accuracy and convergence
properties. It is a reference application for supercomputer benchmarking,
thanks to its good scaling capabilities.

%%% SPECFEM3D %%%
\subsubsection{SPECFEM3D}

When we started to work on the project (version v2.1 of July 2013) it
supported graphics card GPU acceleration through NVidia CUDA. This
GPU support came in addition to the MPI support implemented to enable
multi-CPU parallel computing. Most of SPECFEM3D code-base is written in
Fortran 2003 and only the GPU related parts are written in
C.  The split between CPU and GPU code was done at a rather fine
grain, as the application counted more than 40 GPU kernels. Some of
them were quite simple (\eg performing a few vector operations, but at
massively parallel scale), while at the other side of the spectrum,
some complex kernels took more than 80 parameters and perform very
specific physical transformations.

@@

%%%
\paragraph{Porting kernels to BOAST}  Nvidia CUDA and OpenCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
SPECFEM3D internal physics.

@@

%%%
\paragraph{Evaluation} Our OpenCL/BOAST port of SPECFEM3D is now
merged in SPECFEM3D's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
CUDA version and our BOAST-CUDA version. With the same set of
optimization flags, BOAST CUDA and OpenCL version reported similar
execution time spans. The best execution speed was achieved with CUDA
version though (25\% higher than OpenCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed to OpenCL run-time, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in OpenCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in CUDA. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in OpenCL.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.8 (from 7500 to less than 4000 LOC, partly
because of code duplication, also removing manually unrolled
loops). This is beneficial for SPECFEM3D as it improves the readability
and maintainability of its source code.

@@

In the same mindset, we provided SPECFEM3D test-suite with kernel
performance evaluation mechanisms. These tests will help developers to
try new optimizations in kernel code and measure their impact, without
executing the whole application.

>>>>

%%% Porting SPECFEM3D to OpenCL using BOAST %%%
\subsection{Porting Application Kernels: Porting SPECFEM3D from CUDA to
OpenCL using BOAST\label{subsec:specfem}}

In this last subsection, we present an alternative use-case of the BOAST
framework, focused on portability. As part of the Mont-Blanc project,
we had to port a scientific application so that it could be used to
benchmark the Mont-Blanc HPC cluster. It was composed of 40 complex
and hand-tuned CUDA kernels, and we had to port them to OpenCL. This
scenario also highlights the factorization and maintainability
improvements of the BOAST framework, as we obtained a single BOAST
implementation of the kernels (instead of CUDA plus OpenCL) and this
version of the code is still used and updated in the upstream project.

The challenges we faced in this case-study were more related to the
complexity of the scientific application and its kernels than to the
BOAST framework.
%%% SPECFEM3D %%%
\subsubsection{SPECFEM3D}

SPECFEM3D GLOBE\footnote{SPECFEM3D GLOBE --- CIG
  \url{http://www.geodynamics.org/cig/software/specfem3d-globe}} is a
free seismic wave propagation simulator. It simulates seismic wave
propagation at the local or regional scale based upon spectral-element
method (SEM), with very good accuracy and convergence properties. It
is a reference application for supercomputer benchmarking, thanks to
its good scaling capabilities.

When we started to work on the project (version v2.1 of July 2013) it
supported graphics card GPU acceleration through NVidia CUDA. This
GPU support came in addition to the MPI support implemented to enable
multi-CPU parallel computing. Most of SPECFEM3D code-base is written in
Fortran 2003 and only the GPU related parts are written in
C.  The split between CPU and GPU code was done at a rather fine
grain, as the application counted more than 40 GPU kernels. Some of
them were quite simple (\eg performing a few vector operations, but at
massively parallel scale), while at the other side of the spectrum,
some complex kernels took more than 80 parameters and performed very
specific physical transformations.

@@

%%%
\paragraph{Porting kernels to BOAST}  NVidia CUDA and OpenCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
SPECFEM3D internal physics.

@@

%%%
\paragraph{Evaluation} Our OpenCL/BOAST port of SPECFEM3D is now
merged in SPECFEM3D's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and 24 Intel
Xeon processors, we measured identical performance between the original
CUDA version and our BOAST-CUDA version. This result was expected, as
the BOAST-generated version is identical to the original, apart from
naming differences.

With the same set of
optimization flags, BOAST CUDA and OpenCL versions reported similar
execution time spans. The best execution speed was achieved with CUDA
version though (25\% higher than OpenCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed to the OpenCL run-time, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in OpenCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in CUDA. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in OpenCL.

By refactoring GPU kernels' code in BOAST EDSL, the size of the kernels' source code
shrank by a factor of 1.8 (from 7500 to less than 4000 LOC, mainly
because of the removal of code duplication and manually unrolled
loops). This is beneficial for SPECFEM3D as it improves the readability
and maintainability of its source code. This gain was confirmed by the
SPECFEM3D community as the subsequent developments have been carried out
on the BOAST version of the kernels, and not on the CUDA or OpenCL generated code.

@@

In the same mindset, we provided SPECFEM3D test-suite with kernel
performance evaluation mechanisms. These tests will help developers to
try new optimizations in kernels' code and measure their impact,
without executing the whole application.
